---
title: "Final Project R: João Teixeira"
output: pdf_document
date: "2023-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, include=FALSE}
chooseCRANmirror(graphics=FALSE, ind=1) 

suppressMessages(install.packages(c("Rcpp", "forcats", "reshape2", "ggplot2", "dplyr", "tidyr", dependencies = TRUE)))

library(dplyr)
library(tidyr)
library(ggplot2)
library(Rcpp)
library(reshape2)
library(ggplot2)
library(forcats)

df <- read.csv("https://raw.githubusercontent.com/joaoteixeira1996/R_Final_Assignment-/main/Energy_Transition%20(3).csv")

# Relabel the levels of the Technology variable
df$Technology <- fct_recode(df$Technology,
  "Hydropower" = "Hydropower (excl. Pumped Storage)",
  "Other non-renewable" = "Other non-renewable energy",
  "Geothermal" = "Geothermal energy",
  "Marine" = "Marine energy",
  "Solar" = "Solar energy",
  "Wind" = "Wind energy",
  "Fossil" = "Fossil fuels"
)

#getting rid of unnawanted columns
df <- select(df, -ISO2, -ISO3, -Source, -CTS_Name, -CTS_Code, -CTS_Full_Descriptor)

#reshaping
df <- melt(df, id.vars=c('ObjectId', 'Country', 'Indicator', 'Technology', 'Energy_Type', 'Unit'), variable.name='Year')

#taking out the 'F' from the years
df$Year <- as.integer(substring(df$Year, 2))

df <- df %>%
  group_by(Country, Year, Indicator, Technology, Energy_Type, Unit) %>%
  summarize(value = first(value))
 
#reordering
df <- df[, c('Country', 'Year', 'Indicator', 'Technology', 'Energy_Type', 'Unit', 'value')]
```

## R Final Project

## Link to github: https://github.com/joaoteixeira1996/R_Final_Assignment-

### Executive Summary

This project aims to perform data analysis on the evolution of the use of different energy sources around the world. The topic is extremely relavant nowadays, given the impact that energy generation has on global warming. 

To perform this analyzes, I will use the Energy Transition Dataset from the International Renewable Energy Agency, which was accessed from the IMF Climate Change Board. This dataset contains information on the amount of energy produced in different countries from a variety of sources (Solar, Nuclear, etc). 

During the past decades there has been important debates on the matter of energy. First, the debate on climate change has emphasized the importance of moving from fossil fuels to renewable ones. The discussion has been around for many years now, and one should expect to see some results of it on the data. Second, in 2011 the Fukushima nuclear disaster casted doubts on the safety of nuclear power plants, making countries announce the reduction of the production of energy from this source. This dataset allows to visualize these trends on the energy sources and I will do some further investigation on these topics. 

To see the overall picture of energy generation, the next chart is useful.

```{r, echo = FALSE, include=TRUE}

# Subset the data for the world and electricity generation
df_generation_world <- subset(df, Indicator == "Electricity Generation" & Country == 'World')

# Create a stacked bar chart

chart1 <- ggplot(df_generation_world, aes(x=Year, y=value, fill=Technology)) + 
  geom_bar(stat="identity") +
  
  # Add labels and title
  xlab('Year') +
  ylab('Value') +
  ggtitle('Energy Sources | World') +
  
  # Move the legend outside the plot area
  theme(legend.position="bottom",legend.key.height = unit(0.5, "cm"), 
        legend.key.width = unit(0.5, "cm"))
print(chart1)

ggsave("plots/chart1.png")
```

The chart shows some interesting trends that I will explore throughout the project. First, there is an upward trend in energy generation in the world. Fossil fuels represent the biggest chunk of the energy generated and this production has grown in the past two decades. The good news for the environment is the increase in the production of renewable energy such as hydropower, solar, and wind. 

In the next sections I will analyze more carefully some of these trends. 

### Summary of the raw data set 

The raw dataset is on the wide format and has 11 variables available for the period from 2000 to 2020.  The format of it is not really suitable for data analysis, and, therefore, I will need to conduct some data cleaning, which will be explained in the next section. 

Displaying the structure the dataset for an overview of it:

```{r, echo = FALSE, include=TRUE}

df2 <- read.csv("C:/Users/Usuario/Downloads/Energy_Transition (3).csv")

str(df2)

```

The raw data has 33 variables, however 21 of them represent years (indicating the dates) and not real variables. Also, there are seveeral variables indicating some international standardization, such as ISO2 and ISO3, or CTS_Name. These variables are not useful for this project. Therefore, I will remove them. 

Once cleaned, the goal is to prepare a dataset that includes information on the energy sources produced in a given country. For instance, the dataset will allow checking the amount of energy produced from various sources like Bioenergy, Fossil Fuels, Geothermal, etc. in Austria in a particular year

Also, some reshaping will be needed, since the date variables (years) are in the columns and have an 'F' in front of it, such as F2000 indicating the year 2000. 

In summary, data cleaning is required to work on this dataset. The objective is to produce a useful dataset that will allow me to perform the analysis for this project. 
### Data cleaning

As mention in the previous section, the raw data requires several steps of data cleaning. Overall, the data cleaning work on this project consists of getting rid of useless variables for our analysis; reshaping the dataset; converting the string of years into integers and taking the 'F's out of it; grouping the data; and finally reordering to a nicer way of seeing it. 
The data, when downloaded from the IMF portal, comes in the wide format. Besides that, there are many useless information for our project. Finally, the years come in a strange way, with an 'F' in front of it, such as F2023 standing for the year 2023. I will give the details on the data cleaning process now:

The first step was to get rid of unwanted columns, those that would not bring any valuable information to our analysis. 

```{r, eval = FALSE}

df <- select(df, -ISO2, -ISO3, -Source, -CTS_Name, -CTS_Code, -CTS_Full_Descriptor)

```

The next step is to reshape. The dataframe comes in the wide format, which is not ideal for data analysis. Therefore, we want to reshape for the long format. To do so, we gonna use the melt function. With this function, we set the column names of the now reshaped dataframe.There is also a new columns named 'Year' that will be filled in the next step.

```{r, eval = FALSE}

df <- melt(df, id.vars=c('ObjectId', 'Country', 'Indicator', 'Technology', 'Energy_Type', 'Unit'), variable.name='Year')


```

Now, we need to convert the data variable, years, from string to integer in order to store them as the years. They ears in the original dataset comes with an 'F' in from of it, such as F2000. We need to get rig of these 'F's. 

```{r, eval = FALSE}

df$Year <- as.integer(substring(df$Year, 2))

```

The next step is to group the data 

```{r, eval = FALSE}

df <- df %>%
  group_by(Country, Year, Indicator, Technology, Energy_Type, Unit) %>%
  summarize(value = first(value))
```

Finally, for visual reasons, we need to reorder the dataset

```{r, eval = FALSE}

df <- df[, c('Country', 'Year', 'Indicator', 'Technology', 'Energy_Type', 'Unit', 'value')]

```

Now, we can finally visualize the cleaned data from which I will work on 

```{r, echo = FALSE, include=TRUE}

head(df)

```

### Data Exploration

The dataset gives the opportunity to explore a variety of relevant questions regarding energy production around the world on its different sources. 

#### Energy Production Around the World


The first analysis I will perform will take advantage of the fact that the data covers the whole world. Besides having data for individual countries, it is also provided the data aggregated by continents. From this information, interesting insights can be taken. 

```{r, echo = FALSE, include=TRUE}

# Subset the data for the relevant indicators and year
df_filtered <- subset(df, Indicator == "Electricity Generation" & Year == 2020 & Country %in% c('Africa', 'Asia', 'Europe', 'Latin America and the Caribbean', 'Northern America','Oceania'))

# Rename 'Latin America and the Caribbean' to 'LATAM'
df_filtered$Country[df_filtered$Country == 'Latin America and the Caribbean'] <- 'LATAM'

# Rename 'Northern America' to 'N.America'
df_filtered$Country[df_filtered$Country == 'Northern America'] <- 'N.America'

# Create the stacked bar chart
chart2 <- ggplot(df_filtered, aes(x = Country, y = value, fill = Technology)) + 
  geom_bar(stat = "identity") +
  
  # Add labels and title
  xlab('Continent') +
  ylab('Value') +
  ggtitle('Energy Sources per Continents | 2020') +

  # Rename the labels for 'LATAM' and 'N.America'
  scale_x_discrete(labels = function(x) ifelse(x == "LATAM", "LATAM", ifelse(x == "Northern America", "N.America", x))) +
  
  # Move the legend outside the plot area
  theme(legend.position="bottom")+
  guides(fill = guide_legend(nrow = 3))

print(chart2)
ggsave("plots/chart2.png")
```

A few points come from this chart. First, one can verify that Asia is the biggest producer of energy in the world. This should not come as a surprise, given that the continent has the biggest population in the world and some of the biggest producers of energy of the planet such as the Gulf Countries or Central Asia. Europe and North America seem to be almost tied to the second position. 

There seems to be some heterogeneity in the energy sources as well. For instance, fossil fuels seem to take a much bigger part of the energy production in Asia than in Europe or North America. Additionally, Nuclear energy seems to be a relevant source of energy only in Asia, Europe, And North America. In other continents, it does not seem to play a relevant role. 


#### Nuclear Energy 

Nuclear energy is a controversial source of energy. While some advocate it can help combating climate change for being a cleaner source than fossil fuels, many fear its associated risks. And there are reasons for that: in the past decades, some nuclear disasters worried the world, with Chernobyl being the most famous. 

In recent years another accident caused fear, the Fukushima Disaster. Happening in 2011, after that many countries announced they would reduce their production of nuclear energy. With this dataset, we can check if these announcements were really implemented and caused some reduction in the production of nuclear energy.

```{r, echo = FALSE, include=TRUE, fig.width=7, fig.align="center"}

df_nuclear <- subset(df, Indicator == "Electricity Generation" & Technology== "Nuclear" & Country %in% c('World'))


# Create the stacked bar chart
chart3 <- ggplot(df_nuclear, aes(x = Year, y = value, color = Country)) + 
  geom_line() +
  
  # Add labels and title
  xlab('World') +
  ylab('Value') +
  ggtitle('Nuclear Energy') +
  
  
  
  # Move the legend outside the plot area
  theme(legend.position="bottom")
print(chart3)
ggsave("plots/chart3.png")
```

There is a clear fall in the nuclear energy production in 2011, following the Japanese disaster. Of course one cannot take causality just by looking at a chart, but given our knowledge of the reduction in the use of this source of energy, it seems it caused some impact. However, in the following years, the production seem to catch up to its former level. 
 
### Wind x Solar


The final point of this project will be a simple econometirc analysis of the relation between the production of Wind and Solar energy. 

One can imagine that countries that are interested in investing in renewable energies would not focus in only one source, for instance just wind, but in multiple ones. However, it might happen that a country very suitable for wind energy generation, might not be for solar. Nordic countries, due to their cold weather, could be an example of that. Maybe because of their long winter, solar energy is not much popular there are wind might be.

To check what actually happens I will run and exlain the following code.

```{r, echo = FALSE, include=TRUE}
# I had to read the file again because I was having some weird errors when trying to export to PDF
df <- read.csv("https://raw.githubusercontent.com/joaoteixeira1996/R_Final_Assignment-/main/Energy_Transition%20(3).csv")

#getting rid of unnawanted columns
df <- select(df, -ISO2, -ISO3, -Source, -CTS_Name, -CTS_Code, -CTS_Full_Descriptor)

#reshaping
df <- melt(df, id.vars=c('ObjectId', 'Country', 'Indicator', 'Technology', 'Energy_Type', 'Unit'), variable.name='Year')

#taking out the 'F' from the years
df$Year <- as.integer(substring(df$Year, 2))

df <- df %>%
  group_by(Country, Year, Indicator, Technology, Energy_Type, Unit) %>%
  summarize(value = first(value))

```

```{r, echo = TRUE, include=TRUE}
#I start by filtering the dataset, having only the needed Energy
#Technologies (wind and solar) for the last year of available data (2020).
#Also, I need to get rid of countries aggregates, such as continents, G20,
#or World. I do that because I want to deal with countries and not their
#aggregates

subset_df <- df %>% 
  filter(Technology %in% c("Wind energy", "Solar energy"), 
         Year == 2020, 
         !Country %in% c('Advanced Economies', 'Africa', 'Americas', 'Asia', 
                         'Eastern Asia', 'Eastern Europe', 
                         'Emerging and Developing Economies', 'Europe', 'G20', 'G7', 
                         'Latin America and the Caribbean', 'Northern America', 
                         'Northern Europe', 'Southern Asia', 'Southern Europe', 
                         'Sub-Saharan Africa', 'Western Asia', 'Western Europe', 'World',
                         'Central Asia', 'Oceania', 'Oceania sub-regions', 'South-eastern Asia', 'Northern Africa'))

#I reshaped the dataframe to becase easier to create the charts
reshaped_df2 <- subset_df %>%
  select(Country, Year, Technology, Unit, value) %>%
  pivot_wider(names_from = Technology, values_from = value, names_prefix = "", 
              values_fn = list(Value = sum))

#Since I was having problems in Markdown when generating the PDF, I did the
#needed transformation of the solar and wind energy 
# to logs here. I could do that inside the ggplot function, but I was
#getting error messages. Also, I was having problems with the variables
#named with two names (Solar Energy and Wind energy). Apparently R does not
#like spaces in the name of the variables, so I had to rename them 
 
reshaped_df2 <- reshaped_df2 %>% 
  rename(Solar= 'Solar energy', Wind = 'Wind energy') %>%
  mutate(Solar_log = log(Solar),
       Wind_log = log(Wind))

#I get the top 70 producers of solar energy. This will be needed in the next
#steps
top_solar <- reshaped_df2 %>%
  arrange(desc(Solar_log)) %>%
  head(70)

#I get the top 70 producers of wind energy. This will be needed in the next
#steps
top_wind <- reshaped_df2 %>%
  arrange(desc(Wind_log)) %>%
  head(70)

# If I want to create a scatter plot, I need to have both wind and solar
#observations for a given country. Therefore, I will get those countries #that are part of both
# the top producers of wind and solar energy. By doing this, I know a #country has observations for solar and wind production and I will be able #to generate a plot 

common_countries <- intersect(top_wind$Country, top_solar$Country)

# Creating a dataframe only with only these common countries
reshaped_df2 <- reshaped_df2 %>% 
  filter(Country %in% common_countries)

#I can finally make the plot. I will use ggplot. It takes the log of solar
#energy production in the x-axis and the log of wind energy in the y-axis. #geom_points makes it a scatter plot. I also add a linear regression line to #it with geom_smooth. lm indicates it should be a linear model and 
#SE = FALSE indicates there should be no standard errors displayed
  
ggplot(reshaped_df2, aes(x = Solar_log, y = Wind_log)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE) 


ggsave("plots/my_plot4.png")

# I also run a regression, having solar energy as the dependent variable and wind as the independent. 
model <- lm(Solar_log ~ Wind_log, data = reshaped_df2)

summary(model)

```

The scatter plot shows that indeed there is a positive correlation between solar and wind energy production. On average, a country that produces more solar energy also does prodce more wind energy.

The model also confirms this. We can see that the coefficient of 0.51 indicates that a one percent increase in the production of wind energy should generate a 0.51% change in the production of solar energy. The results are highly significant, with a very low p-value. 


### Conclusion

This project aimed to analyse some trend and facts of the world energy production. The relevancy of this topic is evident given the current urgency of solving the climate crisis. 

In order to perform the investigations, I used the concepts learned from our R class. It was an interesting and challenging way of getting hands-on experience with data analysis. 

The project consisted of finding an interesting and useful dataset, which I did by surfing on the IMF website. After that, I cleaned and organized it, as explained in this file. Then, I generated charts and run a simple econometrics model to investigate the relation of the production of wind and solar energy. I find positive and significant relation between these variables. 


